Evolution of Spectral Distributions in Deep-Water Constant Vorticity Flows

### 2019-12-19 Writeup/random_vorticity.tex

\title{Evolution of Narrow-Band Spectra in Deep-Water Constant Vorticity Flows}
\author{Christopher W. Curtis and Mackensie Murphy}
\date{}
\begin{document}
\maketitle
\section*{Introduction}
There is now a relatively wide range of literature which shows that nonlinear instabilities, in particular the modulational instability (MI), are responsible for significant modifications to the statistical properties of water waves; see \cite{dysthe2,dysthe3,onorato,thomas2012nonlinear} among others.   These results have significant impact on the understanding of deep-water rogue wave formation among other oceanographic phenomena.  However, much of our understanding around MI relies on highly idealized assumptions which cannot be expected to hold in natural settings.  In particular, most understandings of MI rely on looking at the stability of perturbations to carrier waves at essentially a fixed wave number.  To address this shortcoming, in a now seminal paper, \cite{alber} analytically studies families of perturbations of wave packets with narrow, but non-zero, spectral width around a central wave number.  An analytic critierion determining when modulational instability (MI) is either manifested or supressed depending on this spectral width is derived.  The results in \cite{alber} were confirmed numerically in \cite{dysthe2,dysthe3} by examing the mean properties associated with ensembles of initial conditions.

However, many physical effects were ignored in the derivation and verification of the results described above. In particular, vorticity was not taken into account, thereby ignoring a central mechanism for the transfer of energy across length scales in oceanic flows.  Thus, building on the work in \cite{thomas2012nonlinear} and \cite{curtis8}, we numerically explore the analytic stability condition derived in \cite{alber}, focusing in particular on the impact that vorticity plays in controlling the effective width for the onset of MI.  Moving beyond just examining the properties of the NLS equation with vorticity, we look at the statistical properties of solutions to a higher-order model, the vor-Dysthe equation derived in \cite{curtis8}.  As shown in \cite{onorato}, the higher-order terms associated with the Dysthe equation can have significant impacts on the statistics of the waves, so it is a nontrivial question to examine how the vorticity will interact with these higher-order nonlinearities.

\section*{Modeling and Stability Theory}
We examine the unsteady nonlinear wave propagation over a constant shear current.  To do this, we assume the fluid velocity has the form
\[
{\bf u} = u({\bf x},t)\hat{{\bf x}} + w({\bf x},t)\hat{{\bf z}} = \omega z\hat{{\bf x}} + \nabla \phi,
\]
where $\phi$ is a harmonic function.  We restrict fluid motion to the $(x,z)$-plane, thereby ignoring transverse variations in the $y$ dimension.   Following standard arguments, e.g.~see \cite{ashton}, the dynamics of the fluid can be determined by solving the free boundary value problem
\begin{align}
\Delta \phi = 0, & & -\infty< z < \eta(x,t), \nonumber \\
\eta_{t} +\left(\omega\eta+\phi_{x}\right)\eta_{x} - \phi_{z} = 0, & & z=\eta(x,t), \label{kine}\\
\phi_{t} + \omega\pd_{x}^{-1}\eta_{t} + \frac{1}{2} \left|{\bf
    u}\right|^{2} + g\eta -
\frac{\sigma}{\rho}\pd_{x}\frac{\eta_{x}}{\sqrt{1+\eta_{x}^{2}}}= 0, &
& z=\eta(x,t), \label{bernou}\\
\lim_{z\rightarrow -\infty} \phi_{z} = 0, \nonumber
\end{align}
where $\eta$ represents the free surface displacement, and $g$, $\rho$, and $\sigma$ represent the acceleration due to gravity, the fluid density, and the coefficient of surface tension respectively.

By choosing a characteristic wave height $a$ and wave length $L$, all quantities can be non-dimensionalized via
\begin{align*}
\tilde{x} = \frac{x}{L}, ~\tilde{z} = \frac{z}{L}, ~ \tilde{t} = \sqrt{\frac{g}{L}}~t, ~\omega = \sqrt{\frac{g}{L}}~\tilde{\omega}, \\
\eta = a \tilde{\eta}, ~ \phi  = a\sqrt{gL}~\tilde{\phi} , ~ \epsilon = \frac{a}{L}.
\end{align*}
By following the AFM approach described in \cite{afm,ashton} and dropping the tildes, the kinematic boundary condition, Equation \eqref{kine}, can be written in terms of surface variables alone via the integro-differential equation
\begin{equation}
\int_{\mathbb{R}}dx~e^{-ikx}e^{\epsilon|k|\eta}\left(\eta_{t} +
  \epsilon \omega \eta \eta_{x} + i \mbox{sgn}(k)Q \right) = 0, ~
k\neq 0,
\label{kine2}
\end{equation}
where $Q=q_{x}$, $q(x,t) = \phi(x,\eta(x,t),t)$.  By integrating over $\mathbb{R}$, we are assuming that both $\eta$ and $Q$ decay to zero sufficiently rapidly in the far field.  We can also readily derive a nearly identical expression on domains periodic in the horizontal variable $x$, in which case Equation \eqref{kine2} becomes
\begin{equation}
\int_{-L_{p}/2}^{L_{p}/2}dx~e^{-ikx}e^{\epsilon|k|\eta}\left(\eta_{t} +
  \epsilon \omega \eta \eta_{x} + i \mbox{sgn}(k)Q \right) = 0, ~ k = \frac{2\pi m}{L_{p}}, ~ m\in \mathbb{Z}\backslash{0},
\label{integro1per}
\end{equation}
where $L_p$ is the spatial period.  Throughout the remainder of this section we only present results over the real line, $\mathbb{R}$, since identical results can be derived for the periodic case.
Lastly, we note that if in Equation \eqref{kine2} we approach $k=0$ from both the left and the right, we get the equations
\[
\int_{\mathbb{R}}dx \left(\eta_{t} + i Q \right) = 0, ~ \int_{\mathbb{R}}dx \left(\eta_{t} - i Q \right) = 0.
\]
Thus we get the identities
\begin{align}
\pd_{t}\int_{\mathbb{R}} dx ~\eta(x,t) & = 0, \label{zavsurf}\\
\int_{\mathbb{R}} dx ~Q(x,t) & = 0. \label{zavpot}
\end{align}

Taylor expanding Equation (\ref{kine2}) up to $\mathcal{O}(\epsilon^3)$ for $k\neq 0$ gives
\begin{multline}
\int_{\mathbb{R}}dx~e^{-ikx}\left(1 + \epsilon |k|\eta + \frac{\epsilon^{2}|k|^{2}\eta^{2}}{2} \right)\left(\eta_{t} + i \mbox{sgn}(k)Q \right) \\
+ \epsilon \omega \int_{\mathbb{R}}dx~e^{-ikx}\left(1 + \epsilon |k|\eta \right) \eta \eta_{x} = 0.
\label{integro1}
\end{multline}
Transforming into surface variables and Taylor expanding Equation \eqref{bernou}, Bernoulli's equation, up to $\mathcal{O}(\epsilon^3)$ gives
\begin{multline}
Q_{t} + \omega \eta_{t} + \eta_{x} - \tilde{\sigma}\eta_{xxx} + \frac{\epsilon}{2}\pd_{x}\left(-\eta_{t}^{2} + (Q+\omega \eta)^{2} \right)\\
+ \epsilon^{2}\pd_{x}\left( \frac{3}{2}\tilde{\sigma}\eta_{x}^{2}\eta_{xx} - \eta_{t}\eta_{x} \left(Q  + \omega \eta \right) \right)  = 0,
\label{berexp}
\end{multline}
where the reciprocal of the Bond number, $\tilde{\sigma}$, is given by
\[
\tilde{\sigma} = \frac{\sigma}{\rho g L^2}.
\]
We note that we have tacitly assumed $\omega = \mathcal{O}(1)$.  In physical terms, this implies that $\omega$ is comparable to the natural time scale of this problem, $\sqrt{L/g}$.  If we were to assume $\omega$ were of larger magnitude, the problem would no longer be weakly nonlinear and would be much less amenable to asymptotic analysis.  Therefore, throughout the remainder of the paper, we assume that the vorticity is not too large.

By expanding Equation \eqref{integro1} and taking inverse Fourier transforms, we can derive the first three terms of the corresponding Dirichlet-to-Neumann Operator (DNO) expansion, so that we find
\begin{equation}
\eta_{t} = G_{0}Q + \epsilon G_{1}Q + \epsilon^{2}G_{2}Q,
\label{dnoexp}
\end{equation}
where
\begin{align*}
G_{0}Q = & -\mathcal{H}Q\\
G_{1}Q = & \mathcal{H}\pd_{x}\left(\eta G_{0}Q\right) - \pd_{x}\left(\eta Q + \frac{\omega}{2}\eta^{2}\right)\\
G_{2}Q = & \mathcal{H}\pd_{x}\left(\eta G_{1}Q\right) + \frac{1}{2}\pd_{x}^{2}\left(\eta^{2}G_{0}Q \right) + \pd_{x}^{2}\mathcal{H}\left(\frac{1}{2}\eta^{2}Q + \frac{\omega}{3}\eta^{3}\right), 
\end{align*}
where $\mathcal{H}$ denotes the Hilbert transform so that 
\[
\mathcal{H}f = \frac{1}{2\pi}\int_{\mathbb{R}}dk~ e^{ikx}  i\mbox{sgn}(k) \hat{f}(k),
\]
where $\hat{f}(k)$ is the Fourier transform of $f(x)$ defined by
\[
\hat{f}(k) = \int_{\mathbb{R}}dx~ e^{-ikx}f(x).
\]
Using the DNO expansion in Equation \eqref{dnoexp}, we can rewrite Equation \eqref{berexp} an an evolution equation for $Q_{t}$, thereby giving us a closed system in terms of surface variables alone.  The system of equations given by Equations \eqref{dnoexp} and \eqref{berexp} we denote as the Full-Euler Equations (FEEs).

As is well known by now, in non-dimensional variables, we can model the evolution of a relatively small free-fluid surface over infinitely-deep water $z=\epsilon \eta(x,t)$ via the expansion 
\begin{multline*}
\eta(x,t) = \eta_{1}(\xi,\tau)e^{i\theta(x,t)} + \eta^{\ast}_{1}(\xi,\tau)e^{-i\theta(x,t)}\\
+ \epsilon\left(\eta_{0}(\xi,\tau) + \eta_{2}(\xi,\tau)e^{2i\theta(x,t)} + \eta^{\ast}_{2}(\xi,\tau)e^{-2i\theta(x,t)}\right) + \mathcal{O}\left(\epsilon^{2} \right)
\end{multline*}
where $\tau = \epsilon^{2}t$, $\xi = \epsilon(x+c_{g}t)$, and $\theta(x,t) = k_{0}x + \Omega(k_{0},\omega)t$, where the linear dispersion relationship $\Omega(k_{0},\omega)$ is given by
\[
\Omega(k_{0},\omega) = \frac{1}{2}\left(s\omega \pm \sqrt{\omega^{2} + 4|k_{0}|\left(1+\tilde{\sigma}k_{0}^{2}\right)} \right).
\]
Here, $s = \mbox{sgn}(k_{0})$, $\omega$ is the non-dimensionalized magnitude of the vorticity of the flow, and $\tilde{\sigma}$ is the surface tension.

To describe the evolution of the slowly evolving envelope $\eta_{1}(\xi,\tau)$, we use two asymptotically accurate models.  The first model is the the nonlinear Schr\"{o}dinger equation (NLSE) derived in \cite{curtis8}, though see also \cite{thomas2012nonlinear}, given by 
\[
i\pd_{\tau}\eta_{1} + \alpha_{d}\pd_{\xi}^{2}\eta_{1} + \alpha_{nl}\left|\eta_{1} \right|^{2}\eta_{1} = 0, 
\]
where 
\begin{align*}
c_{g} = & \frac{1+3\tilde{\sigma}k_{0}^{2}}{2s\Omega - \omega},\\
\alpha_{d}(k_{0},\omega) = & \frac{(c^2_{g} - 3|k_{0}|\tilde{\sigma})}{2\Omega-s\omega},\\
\alpha_{nl}(k_{0},\omega) = & \frac{k_{0}\left( sk_{0}^{3}\left(8 + \tilde{\sigma}k_{0}^{2} + 2(\tilde{\sigma}k_{0}^{2})^{2}\right) + \omega \alpha_{v}\right)}{\left(2s\Omega -\omega\right)(1+c_{g}\omega)\left(4\Omega^2-s(2k_{0}(1+4\tilde{\sigma}k_{0}^{2})+2\omega\Omega)\right)}.
\end{align*}

To generate simulations, we start both the NLSE and FEE from the same collection of $2L$ periodic-initial conditions
\[
\eta_{1}(\xi,0) = \epsilon_{rms}\sqrt{\frac{\delta \tilde{k}}{\sigma\sqrt{\pi}}} \sum_{k=-K+1}^{K}e^{-\tilde{k}^{2}/2\sigma^{2}}e^{i\theta_{k}}e^{i\tilde{k}\xi}, ~ \tilde{k} = \frac{\pi k}{L},
\]
where $\delta \tilde{k}=\pi/L$ and the phases $\theta_{k}$ are randomly chosen uniformly between $0$ and $2\pi$.  We can then readily show that, for $L\gg 1$, 
\[
\epsilon_{rms} \approx \overline{\left|\eta_{1}\right|^{2}}^{1/2},
\]
where $\overline{()}$ denotes the ensemble average.  In our coordinates then, the work in \cite{alber} and later confirmed in \cite{dysthe2} shows that for the NLSE the MI is suppressed for spectral widths $\sigma$ such that 
\[
\sigma \geq \epsilon_{rms}\sqrt{\frac{2\alpha_{nl}}{\alpha_{d}}}
\]
This work has helped motivate defining this parameter as the Benjamin--Feir Index (BFI), which has been shown to be a critical parameter to understanding the statstical properties of nonlinear free-surface flows; see \cite{onorato, thomas2012nonlinear}.
\section*{Results}
To understand the relationship between the vorticity $\omega$ and the initial spectral width $\sigma$ of our initial conditions, we generate ensembles of initial conditions and then run these different initial conditions up to $\tau=1$, which corresponds to $t_{f} = 1/\epsilon^{2}$ in the original fast variables.  The domain is $[-50\pi,50\pi]$.  The number of modes is $K = 256$.  The surface tension is $\tilde{\sigma} = 10^{-5}$. We then plot the averaged spectral density $S(k,t_{f};\sigma)$ where 
\[
S\left(k,t_{f};\sigma\right) = \overline{\left( \left|\frac{1}{\epsilon}\hat{\eta}_{1}\left(\frac{k-k_{0}}{\epsilon},1;\sigma \right)\right|^{2}\right)}.
\]
which corresponds to the autocorrelation of the surface profile with respect to space.  Note we have emphasized the dependence on the initial spectral width $\sigma$ for the sake of clarity.  Following the approach in \cite{dysthe2,dysthe3}, by comparing the initial spectral density to that at later times, we can see directly whether MI manifests as significant distortions of the statistics, or if through supression the distributions remain relatively close to one another.

Throughout our results, we have used a psuedo-spectral method in space coupled with a 4th-order Runge-Kutta scheme using integrating factors in time, with a time step of $dt = 2.5 \times 10^{-3}$.  De-aliasing is implemented via Orszag's `2/3-rule'.  Averaging was done with $N_{e}=3200$ ensemble members, thereby ensuring relative averaging errors on the order of about $1\%$.  Throughout this section, we present comparisons of the distributions for particular choices of $\sigma$ while also comparing the mean and standard deviations of the respective distributions so as to get a more quantitative understanding of the differences between the different distributions.

\subsection*{Strong-Counter Shear: $\omega = -1$}

Motivated by the results in \cite{curtis8}, we anticipate that $\omega=-1$ presents the most case which would allow for the vorticity to have the strongest impact on the BFI condition, which for the parameter choices requires that $\sigma > 11.9$.  This is borne out by numerical experiment as seen in Figures \ref{fig:meanstdomn1} and \ref{fig:pspecomn1}.  In this case, we see that for the NLSE the BFI condition largely holds.  In particular, we see that while the mean never changes in any significant way from that of the initial distribution, increasing the width $\sigma$ causes the standard deviation of the distribution determined via evolution of the NLSE to collapse onto that of the initial, thereby showing that wider spreads in initial conditions are overall less susceptible to MI.

\begin{figure}[!ht]
\centering
\begin{tabular}{c}
\includegraphics[width=.85\textwidth]{omega_n1_ep0pt1_dt5en4}\\
(a) $\epsilon=.1$ \\
\includegraphics[width=.85\textwidth]{omega_n1_ep0pt04}\\
(b) $\epsilon=.04$
\end{tabular}
\caption{For $\omega=-1$, comparison of the means and standard deviations between the initial averaged spectral density and those generated by the NLSE and VDE.  This is shown for varying strengths of the small amplitude, and thus nonlinearity, parameter $\epsilon$ with $\epsilon=.1$ in (a) and $\epsilon=.04$ in (b).  The vertical line denotes the spectral width corresponding to the BFI condition.}
\label{fig:meanstdomn1}
\end{figure}

\begin{figure}[!ht]
\centering
\begin{tabular}{c}
\includegraphics[width=.65\textwidth]{omega_n1_ep0pt1_dt5en4_width12} \\
(a) $\epsilon=.1$\\
\includegraphics[width=.65\textwidth]{omega_n1_ep0pt04_width12}\\
(b) $\epsilon=.04$ 
\end{tabular}
\caption{For $\omega=-1$, comparison of the averaged spectral densities between the initial averaged spectral density and those generated by the NLSE and VDE.  This is shown for varying strengths of the small amplitude, and thus nonlinearity, parameter $\epsilon$ with $\epsilon=.1$ in (a) and $\epsilon=.04$ in (b).}
\label{fig:pspecomn1}
\end{figure}

However, the results for the VDE are markedly different.  As seen in Figure \ref{fig:pspecomn1}, where $\sigma=12$ so that we are just past the BFI threshold, the most peculiar feature of the VDE is the bimodality of the distribution.  There is clearly a frequency downshift of the central initial peak, but a secondary peak at high-wave number forms as well.  This formation is less pronounced with decreasing $\epsilon$, but is clearly there nevertheless.  This represents a kind of resonance condition for which no ready explanation exists. This feature is likewise responsible for the strong skewing of the means and standard deviations seen in Figure \ref{fig:meanstdomn1}.

>>>>

\title{Evolution of Spectral Distributions in Deep-Water Constant Vorticity Flows}
\author{Christopher W. Curtis and Mackensie Murphy}
\date{}
\begin{document}
\maketitle
\section*{Abstract}
A central question in sea-state modeling is the role that various physical effects have on the evolution of the statistical properties of random sea states.  This becomes a critical issue when one is concerned with the likelihood of rare events such as rogue, or freak, waves which can have significant destructive potential on deep sea ships and other offshore structures.  In this paper then, using a recently derived higher-order model of deep water nonlinear waves, we examine the impact of constant vorticity currents on the statistical properties of nonlinearly evolving random sea states.  As we show, these currents can both decrease and increase the kurtosis of the affiliated distributions of the sea states, thereby diminishing or enhancing the likelihood of rare events.   We likewise numerically study the relationship between the kurtosis and a non-dimensional parameter, the Benjamin--Feir Index, which has proven to be a useful measure of when rare events are likely in oceanographic application.

\section{Introduction}
There is now a wide range of literature which shows that nonlinear instabilities, in particular the modulational instability (MI), are responsible for significant modifications to the statistical properties of water waves; see \cite{dysthe2,dysthe3,onorato,thomas2012nonlinear} among others.   As explained in \cite{janssen}, MI is especially important due to the relative rapidity with which it acts, being orders of magnitude faster than say four-wave resonant interactions which form the traditional backbone of our understanding of how nonlinearity drives changes in statistics of random-wave fields \cite{hasselmann,holthuijsen}.  These results have significant impact on the understanding of deep-water-rogue wave formation among other oceanographic phenomena.

However, much of the understanding around MI relies on highly idealized assumptions which cannot be expected to hold in natural settings.  In particular, most understandings of MI rely on looking at the stability of perturbations to carrier waves at essentially a fixed wave number.  To address this shortcoming, in now seminal papers, \cite{alber} and \cite{alber2} analytically studied families of perturbations of wave packets with narrow, but non-zero, spectral width around a central wave number.  Using the Nonlinear Schr\"{o}dinger equation (NLSE), an analytic criterion determining when MI is either manifested or suppressed depending on this spectral width, now referred to as a Benjamin--Feir index (BFI) was derived.  The results in \cite{alber} were explored numerically in \cite{dysthe2,dysthe3} by examining the mean properties associated with ensembles of initial conditions.

As shown in \cite{janssen} though, the movement of eigenvalues and corresponding bifurcation in stability predicted in \cite{alber} appears at best in an ambiguous fashion.  Regardless though, the BFI was still shown in \cite{janssen} to be an especially useful parameter since it allows for a ready prediction of the kurtosis of a random wave field as a quadratic function of the BFI.  Thus, the BFI becomes a useful predictive tool for the appearance of rare events, such as rogue waves in a random-nonlinear-wave field.  Experimental explorations of the results in \cite{janssen} and \cite{alber} were undertaken in \cite{onorato}, where the role of using higher-order nonlinear models than the NLSE were examined as well.

Later work has simulated the impact of wind and damping on the interplay between the BFI and the kurtosis of a wave field; see \cite{slunyaev,eeltink}.  However, the impact of vorticity was not taken into account in these works, thereby ignoring a central mechanism for the transfer of energy across length scales in oceanic flows.  Moreover, as shown recently in \cite{curtis8}, linear shear profiles can have strong impacts on the behavior of mean properties of surface wave flows with some shear profiles being able to suppress the Stokes Drift at the surface.  Therefore while a relatively idealized case of shear, the interplay between nonlinearity and constant vorticity can manifest in markedly different physics.  Building then on the work in \cite{thomas2012nonlinear} and \cite{curtis8}, and complementing the results in \cite{touboul}, we numerically explore the statistical properties of random-nonlinear-wave fields moving over deep water and linear shear profiles.  Moving beyond just examining the properties of the NLSE with vorticity, we look at the statistical properties of solutions to a higher-order model, the vor-Dysthe equation (VDE) derived in \cite{curtis8}.  As shown in \cite{onorato,janssen} and \cite{eeltink}, the higher-order terms associated with the Dysthe equation can have significant impacts on the statistics of the waves, so it is a nontrivial question to examine how the vorticity interacts with these higher-order nonlinearities.

Using our simulations, starting from normally distributed random-wave fields, we show that the BFI condition with respect to predicting stability is not especially useful, conforming to the results in \cite{janssen}.  While we are able to show that for sufficiently wide enough initial distributions that the temporal modifications to the profile are slight, whether or not this corresponds to an actual bifurcation in stability or simply reflects that for a wide enough profile the sideband growth affiliated with the MI has little impact is not clear.  We note though that classic MI analysis does provide a useful prediction of the time scale over which the most significant dynamics in the statistics of the wave field occur.

Beyond the issue of MI, we see several interesting results from our work.  In particular, we see that increasing vorticity enhances the kurtosis of the random-wave field.  Moreover, the VDE always predicts a larger kurtosis than the NLSE, with this difference between the models exacerbated by increasing vorticity.  Likewise the affiliated statistics of the spectral distribution show marked differences between the two models, with the VDE tending to strongly skew distributions away from Gaussian conditions through peak down shifting and one sided tail enhancement.  Finally, we study the behavior of the kurtosis as a function of the BFI parameter.  The quadratic relationship found in \cite{janssen} and \cite{eeltink} is shown to hold, though the nature of the fit is clearly dependent on the vorticity, conforming to the dependence of the fit on physical parameters found in \cite{eeltink}.

 Thus we show that vorticity and higher-order deep water models can have significant impacts on our understanding of the likelihood of rare events, like rogue waves, in deep-water-surface flows.  In particular, shear profiles which tend to propagate in the direction of the surface wave-field acting on relatively narrow distributions of waves can create strongly non-Gaussian distributions with significantly enhanced tails thereby encouraging the occurrence of rare events.  These results motivate further study of currents in random-wave fields, especially as they pertain to the presence of wind and wave dissipation.

The structure of the paper is an explanation of the model and discussion of the BFI in Section 2.  Section 3 presents our numerical results, while Section 4 presents a conclusion.

\section{Modeling and Stability Theory}
We examine the unsteady nonlinear wave propagation over a constant shear current.  To do this, we suppose the fluid velocity is of the form
\[
{\bf u} = u({\bf x},t)\hat{{\bf x}} + w({\bf x},t)\hat{{\bf z}} = \omega z\hat{{\bf x}} + \nabla \phi,
\]
where $\phi$ is a harmonic function.  We restrict fluid motion to the $(x,z)$-plane, thereby ignoring transverse variations in the $y$ dimension.   Using well established arguments, e.g.~see \cite{ashton}, the dynamics of the fluid are found by solving the free boundary value problem
\begin{align}
\Delta \phi = 0, & & -\infty< z < \eta(x,t), \nonumber \\
\eta_{t} +\left(\omega\eta+\phi_{x}\right)\eta_{x} - \phi_{z} = 0, & & z=\eta(x,t), \label{kine}\\
\phi_{t} + \omega\pd_{x}^{-1}\eta_{t} + \frac{1}{2} \left|{\bf
    u}\right|^{2} + g\eta -
\frac{\sigma}{\rho}\pd_{x}\frac{\eta_{x}}{\sqrt{1+\eta_{x}^{2}}}= 0, &
& z=\eta(x,t), \label{bernou}\\
\lim_{z\rightarrow -\infty} \phi_{z} = 0, \nonumber
\end{align}
where $\eta$ represents the free surface displacement, and $g$, $\rho$, and $\sigma$ represent the acceleration due to gravity, the fluid density, and the coefficient of surface tension respectively.  The system is non-dimensionalized via the choices 
\begin{align*}
\tilde{x} = \frac{x}{L_{p}}, ~\tilde{z} = \frac{z}{L_{p}}, ~ \tilde{t} = \sqrt{\frac{g}{L_{p}}}~t, ~\omega = \sqrt{\frac{g}{L_{p}}}~\tilde{\omega}, \\
\eta = a \tilde{\eta}, ~ \phi  = a\sqrt{gL_{p}}~\tilde{\phi} , ~ \epsilon = \frac{a}{L_{p}}.
\end{align*}
Note that we are assuming $\omega = \mathcal{O}(1)$ throughout the remainder of the paper.  In physical terms, this implies that $\omega$ is comparable to the natural time scale of this problem, $\sqrt{L_{p}/g}$.  If we were to assume $\omega$ were of larger magnitude, the problem would no longer be weakly nonlinear and asymptotic analysis would be siginificantly more difficult.  Therefore, throughout the remainder of the paper, we assume that the vorticity is not too large.

For a relatively small free-fluid surface over infinitely-deep water $z=\epsilon \eta(x,t)$, we model a modulated-wave packet via the expansion 
\begin{multline*}
\eta(x,t) = \eta_{1}(\xi,\tau)e^{i\theta(x,t)} + \eta^{\ast}_{1}(\xi,\tau)e^{-i\theta(x,t)}\\
+ \epsilon\left(\eta_{0}(\xi,\tau) + \eta_{2}(\xi,\tau)e^{2i\theta(x,t)} + \eta^{\ast}_{2}(\xi,\tau)e^{-2i\theta(x,t)}\right) + \mathcal{O}\left(\epsilon^{2} \right)
\end{multline*}
where $\tau = \epsilon^{2}t$, $\xi = \epsilon(x+c_{g}t)$, and $\theta(x,t) = k_{0}x + \Omega(k_{0},\omega)t$, where the linear dispersion relationship $\Omega(k_{0},\omega)$ is given by
\[
\Omega(k_{0},\omega) = \frac{1}{2}\left(s\omega \pm \sqrt{\omega^{2} + 4|k_{0}|\left(1+\tilde{\sigma}k_{0}^{2}\right)} \right).
\]
We note that the form for $\phi$ follows directly from this expansion, allowing for everything to be modeled by the terms in the expansion in $\eta$.  Here, $s = \mbox{sgn}(k_{0})$, $\omega$ is the non-dimensionalized magnitude of the vorticity of the flow, and $\tilde{\sigma}$ is the surface tension, or where the reciprocal of the Bond number, $\tilde{\sigma}$, is given by
\[
\tilde{\sigma} = \frac{\sigma}{\rho g L_{p}^2}.
\]
Throughout the remainder of this paper, we take the positive branch $+$ of the dispersion relationship, thereby ensuring that $\Omega > 0$.  Thus, for $k_{0}>0$, the surface waves to leading order propagate to the left.  Thus, if $\omega>0$, then throughout the bulk of the fluid, the shear current propagates with the surface waves, or is {\it co-propagating}, while for $\omega<0$, we say the shear current is {\it counter-propagating}.

To describe the evolution of the slowly evolving envelope $\eta_{1}(\xi,\tau)$, we use two models.  The first model is the the nonlinear Schr\"{o}dinger equation (NLSE) derived in \cite{curtis8}, though see also \cite{thomas2012nonlinear}, given by 
\[
i\pd_{\tau}\eta_{1} + \alpha_{d}\pd_{\xi}^{2}\eta_{1} + \alpha_{nl}\left|\eta_{1} \right|^{2}\eta_{1} = 0, 
\]
where 
\begin{align*}
c_{g} = & \frac{1+3\tilde{\sigma}k_{0}^{2}}{2s\Omega - \omega},\\
\alpha_{d}(k_{0},\omega) = & \frac{(c^2_{g} - 3|k_{0}|\tilde{\sigma})}{2\Omega-s\omega},\\
\alpha_{nl}(k_{0},\omega) = & \frac{k_{0}\left( sk_{0}^{3}\left(8 + \tilde{\sigma}k_{0}^{2} + 2(\tilde{\sigma}k_{0}^{2})^{2}\right) + \omega \alpha_{v}\right)}{\left(2s\Omega -\omega\right)(1+c_{g}\omega)\left(4\Omega^2-s(2k_{0}(1+4\tilde{\sigma}k_{0}^{2})+2\omega\Omega)\right)}.
\end{align*}
However, the NLSE is a relatively low-order model known to be only of limited use in real-world modeling situations.  A more robust model, following the work in \cite{dysthe1}, is found by going to the next higher order in $\epsilon$, giving us the equation
\begin{multline}
(\omega-2s\Omega)\pd_{\tau}\eta_{1} + i(c_{g}^{2}s-3k_{0}\tilde{\sigma})\pd_{\xi}^{2}\eta_{1} + 2i\epsilon s c_{g}\pd^{2}_{\xi\tau}\eta_{1} - \epsilon\tilde{\sigma} \pd^{3}_{\xi}\eta_{1} + i\epsilon^{2}c_{h}\pd_{\xi}^{4}\eta_{1} \\
- i \alpha_{nl}(\omega-2s\Omega)\left|\eta_{1}\right|^{2}\eta_{1}
+ \epsilon \tilde{\alpha}_{1}\left|\eta_{1}\right|^{2}\pd_{\xi}\eta_{1}\\
+ \epsilon\tilde{\alpha}_{2}\eta_{1}^{2}\pd_{\xi}\eta^{\ast}_{1} +
\epsilon i\tilde{\alpha}_{3}\eta_{1}\mathcal{H}\pd_{\xi}\left|\eta_{1}\right|^{2}= 0.
\label{dysthe}
\end{multline}
where $\mathcal{H}$ denotes the Hilbert transform so that 
\[
\mathcal{H}f = \frac{1}{2\pi}\int_{\mathbb{R}}dk~ e^{ikx}  i\mbox{sgn}(k) \hat{f}(k),
\]
and where $\hat{f}(k)$ is the Fourier transform of $f(x)$ defined by
\[
\hat{f}(k) = \int_{\mathbb{R}}dx~ e^{-ikx}f(x).
\]

We call Equation \eqref{dysthe} the Vor-Dysthe equation (VDE).  Details of its derivation can be found in \cite{curtis8}, though we note a difference in this work with that in \cite{curtis8}.  In this work, we found the inclusion of the higher-order dispersion associated with the coefficient $c_{h}$, given by
\[
c_{h} = 2sc_{g}(2sc_{g}\alpha_{d}+\tilde{\sigma})/(\omega-2s\Omega) - s\alpha_{d}^{2},
\]
was necessary to remove artificial numerical instabilities.  Similar issues with Dysthe like equations have been noted in \cite{eeltink} among other sources.  Moreover, the coefficients $\tilde{\alpha}_{j}$ are readily derived, but their forms are cumbersome and thus we omit explicitly writing them for the sake of readability.

To generate simulations, we start both the NLSE and VDE from the same collection of $2L$ periodic-initial conditions
\[
\eta_{1}(\xi,0) = \epsilon_{rms}\sqrt{\frac{\delta \tilde{k}}{\sigma\sqrt{\pi}}} \sum_{k=-K+1}^{K}e^{-\tilde{k}^{2}/2\sigma^{2}}e^{i\theta_{k}}e^{i\tilde{k}\xi}, ~ \tilde{k} = \frac{\pi k}{L},
\]
where $\delta \tilde{k}=\pi/L$ and the phases $\theta_{k}$ are randomly and independently chosen uniformly between $0$ and $2\pi$.  We can then readily show that, for $L\gg 1$, 
\[
\epsilon_{rms} \approx \overline{\left|\eta_{1}(\xi,0)\right|^{2}}^{1/2},
\]
where $\overline{()}$ denotes the ensemble average.  In our coordinates then, motivated by the work of \cite{alber} and \cite{dysthe2} and more fully explored in \cite{janssen}, we study the stability of the initial wavepacket distributions with regards to the magnitude of the Benjamin--Feir Index (BFI), where the BFI is defined to be  
\[
\mbox{BFI} = \sqrt{\frac{\alpha_{nl}}{\alpha_{d}}}\frac{\epsilon_{rms}}{\sigma}.
\]
Note, this definition is identical with the spectral width criterion of \cite{alber} and the BFI condition given in \cite{thomas2012nonlinear}; see also \cite{onorato}.  It was claimed in \cite{alber}, and in some sense verified in \cite{dysthe2}, that wave packets should be stable with respect to MI's for $\mbox{BFI}<1$, which corresponds to the initial spectral width $\sigma>\sigma_{c}$, where 
\[
\sigma_{c} = \sqrt{\frac{\alpha_{nl}}{\alpha_{d}}}\epsilon_{rms}.
\]
